---
title: "Open Space Access in California"
subtitle: "Do lower income communities have less access to open space?"
author: "Wade Sedgwick"
format:
  html:
    code-fold: false
    code-summary: "Show code"
editor: visual
execute:
  warning: false
---

The purpose of this blog post is to understand and quantify whether lower-income counties in California have less access to open space. I used American Community Survey (ACS) median-income data from 2018 as well as open space data collected from 2014 to 2018 to perform a linear regression model. Because the data did not fit my question, and I wanted to get community open space, open space data was filtered to exclude all open space greater than 100 acres.

```{r}
#| echo: false
library(tidycensus)
library(tidyverse)
library(ggplot2)
library(tmap)
library(sf)
library(usmap)
library(here)
library(tigris)
library(gridExtra)
```

### Introduction

Open spaces and parks provide people with many physical and mental benefits to their health and wellbeing. Not only do people in close proximity to open space get more exercise and are therefore more active, open spaces provide a community with greenery and vegetation, contributing to improved air quality and lower temperatures. Additionally, closer proximity to open space as a teenager is linked to healthier adults with less risk for serious illnesses. Mentally, being outdoors in green open spaces decreases stress, depression and anxiety. Individuals who are introduced to outdoor spaces as a child were more-likely to feel more connected to nature, and to visit open spaces as an adult.

Studies have shown that lower-income communities are affected negatively by their inaccessibility to open space. I will analyze this disproportional access to open space in California. This analysis will show how lower-income communities are unequally affected in ways that might be deemed as less important, but pose an unknown negative affect on those households. Not only do we as a state and country need to address the disproportionate levels of pollution and lack of access to healthy foods, we also need to address other underlying causes that exacerbate this underlying discrimination. From this analysis, we can begin to actively and equitably understand how to build infrastructure that would give lower-income communities more access to open spaces.

### Data

To answer this question, I compared household median income to percent open space in each county. The geometries were from the 'tigris' package, where counties from filtered states are given. I used the tidycensus package to retrieve data from the American Community Survey (ACS). The data I gathered is a yearly sample of 3 million households from 2018 (the ACS has more accurate decennial country-wide census data as well). This can easily be accessed in Rstudio with the 'get_acs' function and the appropriate key. Additionally, I read in ACS 5 year tract data (2014-2018) on open spaces in California which was downloaded from the California State Parks [Park Access Tool](https://www.parksforcalifornia.org/parkaccess/?overlays1=parks%2Cnoparkaccess&overlays2=parks%2Cparksper1000). This data set has the county FIPS code, open space acreage, and geometries, which I aggregated to get open space acreage by county.

```{r, results='hide'}
ca_2018 <- get_acs(geography = "county", 
              variables = c(medincome = "B19013_001"), 
              state = "CA", 
              year = 2018,
              geometry = TRUE)

ca_2018 <- ca_2018 %>%
  mutate(NAME = str_remove_all(NAME, ", California"))

ca_counties <- counties(state = "CA")

open_space <- st_read(file.path("/Users/wsedgwick/Desktop/bren_meds/BREN MEDS/wsedgwick.github.io/posts/2022-12-06-eds222-final-open-space/data/ACS2018_5Y_Tract_PPK.shp"))
```

### Analysis

My initial visualization involved creating a map of California counties, where the variables median income and percent open space are plotted. This revealed that there was a negative relationship between median income and percent open space. It can be seen that counties in the Bay Area, such as San Francisco, San Mateo, and Santa Clara Counties, have higher median income but lower percent open space. Similarly, Mono and Inyo Counties have low median income but have high percent access to open space.

```{r}
open_space_county <- open_space %>% 
  group_by(COUNTYFP) %>% 
  summarise(total_park_acres = sum(Park_Acres))

open_space_county <- open_space_county %>% 
  st_as_sf() %>% 
  st_transform(crs = 4269)

open_space_county_ng <- st_drop_geometry(open_space_county)

ca_2018_ng <- ca_2018 %>% 
  rename(medincome_2018 = "estimate",
         moe_2018 = "moe") %>% 
  mutate(NAME = str_remove_all(NAME, ", California")) %>% 
  st_drop_geometry()

ca_counties_joined <- ca_counties %>% 
  dplyr::select("COUNTYFP", "GEOID", "NAME", "NAMELSAD", "ALAND", "AWATER") %>% 
  full_join(open_space_county_ng, by = "COUNTYFP") %>%
  full_join(ca_2018_ng, by = "GEOID") %>%
  mutate(percent_open_space = (total_park_acres / ALAND) * 100,
         income_lower = medincome_2018 < 71228)

map_medincome <- tm_shape(ca_2018) +
  tm_polygons(col = "estimate",
              title = "Median Income") +
  tm_layout(main.title = "Median Income\n(ACS 2018)")

map_open_space <- tm_shape(ca_counties_joined) +
  tm_polygons(col = "percent_open_space",
              title = "Percent open space") +
  tm_layout(main.title = "Percent Open Space\n(ACS 2018 5 year Tract)")

tmap_arrange(map_medincome, map_open_space, ncol = 2)
```

After mapping the two variables, I plotted and ran a linear regression model to quantify the relationship. The R-squared and Adjusted R-squared values were 0.082 and 0.066, respectively, so there was little relationship between median income and percent open space with the data I used. However, the p-value is below our significance level of 0.05, which means that median income is a predictor of percent open space.

```{r}
ggplot(data = ca_counties_joined, aes(x = medincome_2018, y = percent_open_space)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Percent Open Space as a function of Median Income (2018)",
       x = "Median Income of California Counties ($)",
       y = "Percent Open Space") +
  lims(x = c(40000, 120000))

mod <- lm(formula = percent_open_space ~ medincome_2018, data = ca_counties_joined)
summary(mod)
```

With these results, the data set did not reflect the analysis I wanted to perform. Because my focus is on open space access to communities, I filtered open spaces that might not be accessible to communities in their daily lives. All open spaces greater than 100 acres, which are less accessible to households in their daily lives, were filtered out. 79,313 acres of open space remained in the data set, while 47.7 million acres were filtered out.
